Promise and peril: Agnostic identification methods\\ for detecting differential item functioning

### 2020-03-21 paper/paper.tex
after knitting revisions from ben d

\begin{document}
\maketitle
\begin{abstract}
It is well known that likelihood ratio tests (LRT) are effective at detecting differential item functioning (DIF) in item response models. However, to use an LRT, a model identifying assumption must be made so that differences in group ability can be disentangled from potential DIF. We use the term ``agnostic identification'' (AI) to describe the process of finding such identifying assumptions without a priori knowledge of relative group ability or items that are DIF-free. We summarize existing AI methods and propose a variety of new methods. We conduct a simulation study --- which we argue is more realistic than most DIF simulation studies in the literature --- and find that one of the proposed new AI methods, All-others-as-anchors-one-at-a-time (AOAA-OAT), significantly outperforms current methods. We also offer a new method, the equal means, multiple imputation logit graph (EM-MILG), that clearly presents all information about possible DIF, including sampling variability in item parameters, to the analyst. \clearpage
\end{abstract}

@@

@@

Following Camilli (\protect\hyperlink{ref-camilli1992conceptual}{1992}), we conceptualize differential item functioning (DIF) as a varying relationship between ability and probability of correct response for students from different groups that manifests when an item response model with too few ability dimensions is imposed. From this perspective, the term \enquote{differential item functioning} is, perhaps, a misnomer as DIF is better thought of as a property of the student, as opposed to the item. For example, Ackerman (\protect\hyperlink{ref-ackerman1992didactic}{1992}) describes a scenario in which a test intends to measure a student's math ability, but performance also depends on their verbal ability. In this case, math ability is the \enquote{target ability,} and verbal ability is the \enquote{nuisance ability.} Fitting a unidimensional item response model to this test results in students with low verbal ability receiving a score systematically lower than their true math ability; therein lies \enquote{DIF}.

Despite this, the usual setup of DIF simulation studies frames DIF as a property of the item. For example, Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulate students as belonging to either a reference or focal group. They fix the item easinesses for the reference group, \({b_j}^{\text{ref}}\), to values from a previous study. They set item easinesses for the focal group to \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for items without DIF and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.6\) for items with DIF, where 0.6 is the magnitude of DIF in logits. They simulate student ability \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. They generate item responses according to the Rasch model, which specifies that the probability of student \(i\) responding correctly to item \(j\) is
\begin{align}
P(y_{ij} = 1 | \theta_i, b_j) = \sigma(\theta_i + b_j)
\end{align}
where \(\sigma(x) = \frac{e^x}{e^x + 1}\) is the standard logistic function.

For every DIF simulation study framed in terms of item parameters that vary across groups, there is a mathematically equivalent setup in which students' abilities are multidimensional. For example, to translate the Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulation from the DIF-as-item-property view to the DIF-as-student-property view, item easiness is set to what was previously \({b_j}^{\text{ref}}\) for all students. Student ability is expanded to two dimensions, the target ability dimension and nuisance ability dimension. The target ability is what was previously just ability where \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. The nuisance ability is set to \({\eta_i}^{\text{ref}} = 0\) for all students in the reference group and \({\eta_i}^{\text{foc}} = -1\) for all students in the focal group. We now need to use a compensatory 2PL model where the slope on target ability \(a_{1j} = 1\) for all items (consistent with the Rasch model) and the slope on nuisance ability \(a_{2j} = 0.6\) for all items with DIF, and \(a_{2j} = 0\) otherwise. Again, 0.6 is the magnitude of DIF in logits. According to the two-dimensional compensatory 2PL model (Thissen and Steinberg \protect\hyperlink{ref-thissen1986taxonomy}{1986}), the probability student \(i\) responds correctly to item \(j\) is
\begin{align}
\text{Pr}(y_{ij} = 1 | \theta_i, \eta_i, a_{1j}, a_{2j}, b_j) = \sigma(a_{1j}\theta_i + a_{2j}\eta_i + b_j).
\end{align}
This translation between views makes explicit that nearly all DIF simulation studies have, perhaps suboptimally, examined the unrealistic scenario in which there is no variation in the nuisance ability for students in the same group.

If we insist on describing simulation conditions from the DIF-as-student-property view, one might wonder the following: Why not fit a multidimensional item response model which describes the data fully instead of looking for bias in a lower dimensional model? Camilli (\protect\hyperlink{ref-camilli1992conceptual}{1992}) tested this idea with the goal of a \enquote{more satisfying description of the secondary abilities} {[}p.~144{]}. He found that the rotational indeterminacy of item response models is challenging to overcome and concluded that \enquote{a priori knowledge of the true factor structure} is necessary {[}p.~144{]}. It's hard to imagine how one would have such knowledge. Therefore, the best approach, which the DIF literature has nearly unanimously taken, is to fit unidimensional item response models and then look for bias manifesting in the item parameters.

@@

Psychometricians have long been in search of the perfect DIF detection method. It has been well demonstrated that the IRT-based likelihood ratio test (LRT) is effective at detecting DIF (Meade and Wright \protect\hyperlink{ref-meade2012solving}{2012}). As a result, we avoid methods like the Mantel-Haenszel procedure (Holland and Thayer \protect\hyperlink{ref-holland1986differential}{1986}), which muddies the waters by moving away from the IRT framework, and has been shown to perform no better --- and in some cases worse --- than IRT-based methods (Swaminathan and Rogers \protect\hyperlink{ref-swaminathan1990detecting}{1990}).

The unsolved problem that we address here is how to link groups in the typical setting in which no a prior assumptions --- about either relative group ability or which items might have DIF --- can be made. That is, we're interested in agnostic identification methods (hereafter referred to as \enquote{AI methods}) so that LRTs can be used to test items for DIF. For simplicity, we focus on the Rasch model to isolate the fundamental issues in DIF detection. And, in search of a coherent framework, we sometimes edit names of existing methods. We recognize that others have done the same (e.g., Kopf, Zeileis, and Strobl \protect\hyperlink{ref-kopf2015framework}{2015}), and that we risk contributing to a proliferation of names.

In this section, we summarize existing AI methods and propose a variety of extensions. We use simple demonstration data to illustrate how these methods work: 10,000 reference group students and 10,000 focal group students taking an eight-item test. Target ability is simulated \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. Nuisance ability is set to \({\eta_i}^{\text{ref}} = 0\) and \({\eta_i}^{\text{foc}} = -1\). The slope on target ability is set to \(a_{1j} = 1\) for all items. The slope on nuisance ability is set to \(a_{2j} = 0.5\) for the last three items (the items with DIF) and \(a_{2j} = 0\) otherwise. We can also, of course, describe these conditions from the DIF-as-item-property view where there is no nuisance ability. Instead, \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for the first five items and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.5\) for the last three items. To be sure, this data is used only for demonstrative purposes. We compare the performance of the AI methods in a more realistic scenario in the \protect\hyperlink{simstudy}{the simulation study}.

In general, we denote the mean ability of the reference and focal group as \(\mu^\text{ref}\) and \(\mu^\text{foc}\), respectively. We follow the common and inconsequential practice of identifying the scale by setting \(\mu^\text{ref} = 0\). The fundamental challenge is to use an AI method to identify the model, which allows for disentagling the estimation of \(\hat\mu^\text{foc}\) from the estimation of \(\hat d = \hat{b_j}^{\text{foc}} - \hat{b_j}^{\text{ref}}\), the difference in item easiness. The most common AI approach is to select a group of anchor items that are assumed to be DIF-free. These anchor items identify the model, thereby allowing for the estimation of \(\hat\mu^\text{foc}\), and the remaining items can be tested for DIF using an LRT.

Computing is done in R (R Core Team \protect\hyperlink{ref-rcore}{2019}), model fitting in the mirt R package (Chalmers \protect\hyperlink{ref-chalmers2012mirt}{2012}), and data wrangling/visualization in the suite of R packages known as the tidyverse (Wickham \protect\hyperlink{ref-tidy}{2017}). Code is available at \emph{(TODO)}.

@@

AI methods typically work like a black box. The analyst puts their item response data in and the black box outputs an identification assumption to be used when fitting the final model. These types of AI methods have their role --- and the main focus of this paper is on them --- but wouldn't it be nice if there was a way for the analyst to visualize the possibility of DIF in their data?

We think so and, in response, propose the \enquote{equal means, multiple imputation logit graph} (EM-MILG), which presents information about potential DIF to the analyst. EM-MILG begins by fitting a unidimensional Rasch model to the data that is identified by arbitrarily setting \(\mu^\text{foc} = 0\). Recall that \(\mu^\text{ref} = 0\), so this implies that the groups have equal mean ability. The result is that all differences in performance --- either from group ability differences or DIF --- manifest in the item easiness difference parameter \(\tilde{d_j}\). The tilde (e.g., \(\tilde{d_j}\)) is used to denote parameters estimated with \(\mu^\text{foc}\) set to \(0\) --- as compared to the hat (e.g., \(\hat{d_j}\)), which is used for parameter estimates from the final model.

@@

Figure \ref{fig:emmilg} shows the EM-MILG for our demonstration data. We cannot state strongly enough that the EM-MILG contains all possible information about the difference in group performance. The analyst might assume that --- because there are five items where the reference group outperforms the focal group by approximately 1 logit and only three items where the difference is 1.5 logits --- items 1-5 are unbiased. These unbiased items are known as anchor items. After anchor items have been selected, the model is refit where the assumption that \(\mu^\text{foc} = 0\) is dropped and replaced with the assumption that the anchor items are DIF-free (i.e., \(d_j = 0\)).

@@

In general, one of our key concerns with traditional AI methods is that they can lull the analyst into a false sense of security. Too often, the analyst chooses a method, implements it, and then proceeds as if the method certainly resulted in a model that reflects some objective reality. A EM-MILG combats this concern by presenting all information clearly to the analyst. In the previous example, the analyst may be wary of their results, having seen how arbitrary it was to conclude that the first five and not the last three items are unbiased. Even when other AI methods are used, the analyst can use the EM-MILG as the first step in order to give them a sense of their item response data. And, of course, the MILG can be used to visualize DIF regardless of how the model is identified. We now move on to more traditional AI methods, which don't require judgment from the analyst.

@@

Meade and Wright (\protect\hyperlink{ref-meade2012solving}{2012}) compared the most commonly used AI methods and unequivocally recommended the all-others-as-anchors (AOAA) method. AOAA tests each item for DIF one at a time using all of the other items as anchors. For example, when testing the first item for DIF, all of the other items are used as anchors. This is done using an LRT that compares the baseline model, where all item parameters are fixed across groups, to the flexible model, where the parameters of the tested item are freed across groups (Thissen, Steinberg, and Wainer \protect\hyperlink{ref-thissen1993detection}{1993}). Then, when testing the second item for DIF, once again all of the other items (including the first item) are used as anchors, and so on. The items for which the flexible model outperforms the baseline model (typically based on a \(\chi^2\) test) are identified as having DIF, and the rest of the items become anchor items. AOAA is implemented in the mirt R package, and is called by passing scheme = \enquote{drop} to the DIF function (drop refers to dropping a single constraint when moving from the constrained to the flexible model).

Edelen et al. (\protect\hyperlink{ref-edelen2006identification}{2006}) used AOAA to look for DIF between the English and Spanish versions of the 21-item Mini-Mental State Examination and found that 10 of the 21 items exhibited DIF. How can they be sure that its those 10 items and not the other 11 items with DIF? We argue that this method cannot tell us with certainty that it is those 10 items that have DIF while the other 11 do not. Implicit in the use of AOAA is the assumption that all items not being tested do not exhibit DIF, which is, of course, impossible. More practically, it is thought that AOAA will perform well if a small minority of items have DIF or the DIF is balanced such that some items are biased against the focal group, while others are biased against the reference group. Undesirably, most applications of AI methods and many simulation studies do not make explicit the assumptions of the AI method (Strobl et al. \protect\hyperlink{ref-strobl2018anchor}{2018}). In this way, psychometricians might benefit from adopting economists' habit of explicitly stating assumptions and debating their plausibility.

Researchers have noticed the circular logic of AOAA, but have mostly described it indirectly by pointing out inflated Type I errors in simulation studies (Stark, Chernyshenko, and Drasgow \protect\hyperlink{ref-stark2006detecting}{2006}). A simple thought experiment illustrates how AOAA fails: Imagine a three item test with a sufficiently large number of students where the first item has DIF, and the other two do not. Using AOAA, all items test positive for DIF. The last two items incorrectly test positive because including the first item in the anchor set causes the group ability difference to be misestimated. This phenomenon of items with real DIF inducing the appearance of DIF in other items was only indirectly discussed in the literature until Andrich and Hagquist (\protect\hyperlink{ref-andrich2012real}{2012}) coined the term \enquote{artificial DIF.}

@@

One way to attempt to counter artificial DIF is with purification through iterative anchor selection. For example, Drasgow (\protect\hyperlink{ref-drasgow1987study}{1987}) started with AOAA, removed items displaying DIF from the anchor set, then repeated the process iteratively --- with items that have been removed from the anchor set allowed to have free parameters across groups in both the baseline and flexible model --- until no more items tested positively. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) named this technique Iterative-backward-AOAA with \enquote{backward} (as in reverse) referring to beginning with the assumption that all items are DIF-free. We find it clearer to refer to this method as all-others-as-anchors-all-significiant (AOAA-AS). Appending \enquote{all-significant} indicates that anchor selection is made iteratively with all items that test positive for DIF being removed from the anchor set. AOAA-AS is implemented in the mirt R package, and is called by passing scheme = \enquote{drop\_sequential} to the DIF function.

AOAA-AS might seem like an improvement, but it doesn't solve a fundamental problem of AOAA: What does one do when all items test positive for DIF? With a sufficient sample size and at least one item with DIF, this will necessarily be the case. In our thought experiment, we get the same result with AOAA-AS as we did with AOAA: All items test positive for DIF, and there are no anchor items. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) encountered precisely this problem in their simulation study and chose to select a single anchor item randomly. Woods (\protect\hyperlink{ref-woods2009empirical}{2009}) suggested a more straightforward, one-step method which uses AOAA and selects the, say, four items that exhibit the least amount of DIF. It's unclear if one should proceed if those four items also test positive for DIF.

@@

We propose an extension of these methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), which, to our knowledge (and surprise), has not previously been explicitly proposed. AOAA-OAT is inspired by Hagquist and Andrich (\protect\hyperlink{ref-hagquist2017recent}{2017}), who, in general, assert that \enquote{items showing DIF initially should not be resolved simultaneously but sequentially} {[}p.~6{]}. Like AOAA-AS, AOAA-OAT starts with AOAA, but only the single item exhibiting the most DIF, based on the \(\chi^2\) test statistic, is removed from the anchor set. The process continues iteratively until no new items display DIF. AOAA-OAT and AOAA-AS are similar in that they are both iterative; the difference is that AOAA-OAT takes the more conservative approach of removing only \emph{one} item in each iteration as opposed to \emph{all} items that test positive for DIF. As a result, we believe that AOAA-OAT is less likely to be \enquote{fooled} by artifical DIF. AOAA-OAT is not currently implemented in the R package mirt.

Applying AOAA-OAT to our thought experiment demonstrates its effectiveness. The initial AOAA removes the real DIF item from the anchor set because it exhibits the most DIF. In the next step, both of the other items test negative for DIF, and we arrive at the correct conclusion. To work, AOAA-OAT has two requirements: First, that at least two items do not have DIF, and second, that the set of items without DIF are more homogeneous than other sets of items.

@@

Bechger and Maris (\protect\hyperlink{ref-bechger2015statistical}{2015}) proposed selecting anchor items by identifying clusters of items that function similarly and then choosing one of those clusters to be the \enquote{anchor cluster.} They pointed out that one way around the unidentifiability issue is to consider only relative item parameters. For each group, the relative easinesses for each pair of items can be stored in the matrix \(\mathbf{R}^{\text{ref}}\) with entries \({R_{xy}}^{\text{ref}} = {b_x}^{\text{ref}} - {b_y}^{\text{ref}}\), where \(x\) and \(y\) both index the items. To be sure, \(\mathbf{R}^{\text{ref}}\) is a square with the number of rows and columns equal to the number of items. The ultimate matrix of interest is \(\Delta \mathbf{R} \equiv \mathbf{R}^{\text{ref}}-\mathbf{R}^{\text{foc}}\) which is the \enquote{differences between groups in the pairwise differences in (easiness) between items} {[}p.~323{]}.

>>>>

\begin{document}
\maketitle
\begin{abstract}
It is well known that likelihood ratio tests (LRT) are effective at detecting differential item functioning (DIF) in item response models. However, to use a LRT, we require an identifying assumption to disentangle differences in group ability from potential DIF. We use the term ``agnostic identification'' (AI) to describe the process of finding such identifying assumptions without a priori knowledge of relative group ability or items that are DIF-free. We first summarize existing AI methods and propose a variety of new methods. We then conduct a simulation study---which we argue is more realistic than most DIF simulation studies in the literature---and find that one of the proposed new AI methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), significantly outperforms current methods. We also suggest a new approach, the equal means, multiple imputation logit graph (EM-MILG), which clearly presents all information about possible DIF, including sampling variability in item parameters, to the analyst. \clearpage
\end{abstract}

@@

\hypertarget{ben-d-notes}{%
\section{Ben D Notes}\label{ben-d-notes}}

\begin{itemize}
\item
  Switch 1 and 2 to theta and eta
\item
  Ben worries that in anchor items in particular having the demonstration data and the the thought experiment is too many threads
\item
  Be clear upfront that there are two ways to identify: anchor items and anchor points (and be sure to define both)
\item
  Explain AOAA-OAT inuition better. Perhaps using an EM-MILG or something? (in reference to the \enquote{least homogeneous in the sense of} part)
\item
  Compare performance across all methods at once, not one at a time
\item
  Show a paneled MILG for every method all at once. How cool.
\item
  About the simulation difmap figure: \enquote{It seems to me like more could be set about this. you might want to note, for example, that it need not be the case in general that we have 6 items \enquote{clustering} around the same location of the scale. you could have items 1-6 spread out in same way as items 7-12. much harder in that case. in fact, it seems like there is a geometry problem here we could attempt to solve. there are some item-easiness configureations (in sense of MILG) that we really are probably in bad shape with. if the item easiness params are uniformly distributed between -1 and 1 for example, we're stuck! maybe we could try to think about how to characterize such cases and write about what to do there?}
\item
  Fix that equation to take into account sample sizes
\end{itemize}

@@

Following Camilli (\protect\hyperlink{ref-camilli1992conceptual}{1992}), we conceptualize differential item functioning (DIF) as a varying relationship between ability and probability of correct response for students from different groups that manifests when one imposes an item response model with too few ability dimensions. From this perspective, the term \enquote{differential item functioning} is, perhaps, a misnomer as DIF is better thought of as a property of the student, as opposed to the item. For example, Ackerman (\protect\hyperlink{ref-ackerman1992didactic}{1992}) describes a scenario in which a test intends to measure a student's math ability, but performance also depends on their verbal ability. In this case, math ability is the \enquote{target ability,} and verbal ability is the \enquote{nuisance ability.} Fitting a unidimensional item response model to this test results in students with low verbal ability receiving a score systematically lower than their true math ability; therein lies \enquote{DIF}.

Despite this, the usual setup of DIF simulation studies frames DIF as a property of the item. For example, Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulate students as belonging to either a reference or focal group. They fix the item easinesses for the reference group, \({b_j}^{\text{ref}}\), to values that they obtained from a previous study. They set item easinesses for the focal group to \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for items without DIF and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.6\) for items with DIF, where 0.6 is the magnitude of DIF in logits. They then simulate student ability \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. Finally, they generate item responses according to the Rasch model, which specifies that the probability of student \(i\) responding correctly to item \(j\) is
\begin{align}
P(y_{ij} = 1 | \theta_i, b_j) = \sigma(\theta_i + b_j) \label{zopf}
\end{align}
where \(\sigma(x) = \frac{e^x}{e^x + 1}\) is the standard logistic function.

For every DIF simulation study framed in terms of item parameters that vary across groups, there is a mathematically equivalent setup in which students' abilities are multidimensional. For example, to translate the Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulation from the DIF-as-item-property to the DIF-as-student-property view, item easiness is set to what was previously \({b_j}^{\text{ref}}\) for all students. Student ability is expanded to two dimensions, the target ability dimension and nuisance ability dimension. The target ability is the same as unidimensional ability above where \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. The nuisance ability takes one of two values: \({\eta_i}^{\text{ref}} = 0\) for students in the reference group and \({\eta_i}^{\text{foc}} = -1\) for students in the focal group. We then reply on a 2PL compensatory 2PL model; slopes on the target ability are \(a_{1j} = 1\) for all items (consistent with the Rasch model) and the slope on nuisance ability is \(a_{2j} = 0.6\) for all items with DIF, and \(a_{2j} = 0\) otherwise. Again, 0.6 is the magnitude of DIF in logits. According to the two-dimensional compensatory 2PL model (Thissen and Steinberg \protect\hyperlink{ref-thissen1986taxonomy}{1986}), the probability that student \(i\) responds correctly to item \(j\) is
\begin{align}
\text{Pr}(y_{ij} = 1 | \theta_i, \eta_i, a_{1j}, a_{2j}, b_j) = \sigma(a_{1j}\theta_i + a_{2j}\eta_i + b_j).
\end{align}
For a choice of student (from the focal or reference group) and item (with DIF or without), this model produces identical probabilities as Eqn \ref{zopf}. This translation between views makes explicit that nearly all DIF simulation studies have, perhaps suboptimally, examined the unrealistic scenario in which there is no variation in the nuisance ability for students in the same group.

If we insist on describing simulation conditions from the DIF-as-student-property view, one might wonder the following: Why not fit a multidimensional item response model which describes the data fully instead of looking for bias in a lower dimensional model? Camilli (\protect\hyperlink{ref-camilli1992conceptual}{1992}) tested this idea with the goal of a \enquote{more satisfying description of the secondary abilities} {[}p.~144{]}. He found that the rotational indeterminacy of item response models is challenging to overcome and concluded that \enquote{a priori knowledge of the true factor structure} is necessary {[}p.~144{]}. It's hard to imagine how one would have such knowledge. Therefore, the best approach, which the DIF literature has nearly unanimously taken, is to fit unidimensional item response models and then look for bias manifesting in the item parameters. This approach depends upon a crucial identifying assumption; this assumption is the target of our investigation.

@@

Psychometricians have long been in search of the perfect method for detecting differential functioning of \emph{individual} items. The IRT-based likelihood ratio test (LRT) is effective at detecting DIF (Meade and Wright \protect\hyperlink{ref-meade2012solving}{2012}). Given both the efficacy or the LRT approach and our preference for methods that existin within the IRT-framework, we do not discuss methods like the Mantel-Haenszel procedure (Holland and Thayer \protect\hyperlink{ref-holland1986differential}{1986}) (which, for example, has been shown to perform no better---and in some cases worse---than IRT-based methods (Swaminathan and Rogers \protect\hyperlink{ref-swaminathan1990detecting}{1990})).

However, use of LRT requires an identifying assumption. The assumption is needed so as to link groups. In the typical setting, no a prior assumptions---about either relative group ability or which items might have DIF---can be made. Thus, we're interested in \emph{agnostic identification} methods (hereafter referred to as \enquote{AI methods}). For simplicity, we focus on the Rasch model to isolate the fundamental issues in \enquote{the AI problem}. Further, so as to offer a coherent framework, we sometimes edit names of existing methods.\footnote{We recognize that others have done the same (e.g., Kopf, Zeileis, and Strobl \protect\hyperlink{ref-kopf2015framework}{2015}), and that we risk contributing to a proliferation of names.}

In this section, we summarize existing AI methods and propose a variety of extensions. We use a (simulated) demonstration dataset to illustrate key features of the different methods.\footnote{This data is used solely for illustrative purposes; we describe an extensive simulation study later in the paper.} The data consists of 10,000 reference group students and 10,000 focal group students taking an eight-item test. Our simulation is similar to our multidimensional alternative to the Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) approach discussed above. Target ability is simulated as \({\theta_i}^{\text{ref}} \sim N(\mu^\text{ref} = 0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(\mu^\text{foc} = -1,1)\) for students in the focal group. Nuisance ability is set to \({\eta_i}^{\text{ref}} = 0\) and \({\eta_i}^{\text{foc}} = -1\). The slope on target ability is set to \(a_{1j} = 1\) for all items. The slope on nuisance ability is set to \(a_{2j} = 0.5\) for the last three items (the items with DIF) and \(a_{2j} = 0\) otherwise. We can also, of course, describe these conditions from the DIF-as-item-property view where there is no nuisance ability. Instead, \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for the first five items and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.5\) for the last three items. To be sure, this data is used only for demonstrative purposes. We compare the performance of the AI methods in a more realistic scenario in the \protect\hyperlink{simstudy}{the simulation study}.

The fundamental challenge is to use an AI method to disentangle estimation of \(\hat\mu^\text{foc} - \mu^\text{ref}\), the difference in mean abilities, from the estimation of \(\hat d = \hat{b_j}^{\text{foc}} - \hat{b_j}^{\text{ref}}\), the difference in item easiness. The most common AI approaches identify a group of anchor items that are assumed to be DIF-free. These anchor items identify the model, thereby allowing for the estimation of \(\hat\mu^\text{foc}\), and the remaining items can be tested for DIF using an LRT.

Throughout this paper, we estimate models using marginal maximum likelihood estimation (Bock and Aitkin \protect\hyperlink{ref-bock1981marginal}{1981}). In addition, we follow the common and inconsequential practice of identifying the scale by setting \(\mu^\text{ref} = 0\).

@@

AI methods typically work like a black box. The analyst puts their item response data in and the black box outputs an identification assumption to be used when fitting subsequent models.

While we will document the performance of these methods momentarily, we first suggest an approach focused on putting analysts in a better position for understanding the scale of the problem in their data. We propose the \enquote{equal means, multiple imputation logit graph} (EM-MILG), which presents information about potential DIF to the analyst. EM-MILG begins by fitting a unidimensional Rasch model to the data that is identified by arbitrarily setting \(\mu^\text{foc} = 0\). Given that we also assume \(\mu^\text{ref} = 0\), this is equivalent to the assumption that the groups have equal mean ability. As a result, all differences in performance---either from group ability differences or DIF---manifest in the item easiness difference parameter \(\tilde{d_j}\). The tilde (e.g., \(\tilde{d_j}\)) is used to denote parameters estimated with \(\mu^\text{foc}\) set to \(0\)---as compared to the hat (e.g., \(\hat{d_j}\)), which is used for parameter estimates from the properly identified model.

@@

Figure \ref{fig:emmilg} shows the EM-MILG for our demonstration data. We emphasize that the EM-MILG contains all possible information about the difference in group performance. The analyst might assume that---because there are five items where the reference group outperforms the focal group by approximately 1 logit and only three items where the difference is 1.5 logits---items 1-5 are unbiased. Indeed, all AI methods rest on the logic that its the minority of items that contain DIF. These unbiased items are known as anchor items; setting anchor items is the most common identification assumption. After anchor items have been selected, the model is refit where the assumption that \(\mu^\text{foc} = 0\) is dropped and replaced with the assumption that the anchor items are DIF-free (i.e., \(d_j = 0\)). Of course, the demonstration data is designed such that there are obvious groups; this will not typically be the case.

@@

In general, one of our key concerns with traditional AI methods is that they can lull the analyst into a false sense of security. The analyst simply chooses a method, implements it, and then proceeds as if the method certainly resulted in a model that reflects some objective reality. Our aim with EM-MILG is to combat this concern by presenting all information clearly to the analyst. In the previous example, the analyst may be wary of their results, having seen how arbitrary it was to conclude that the first five and not the last three items are unbiased. Even when other AI methods are used, the analyst can use the EM-MILG as the first step in order to give them a sense of their item response data. And, of course, the MILG can be used to visualize DIF regardless of how the model is identified. We now move on to more traditional AI methods, which don't require judgment from the analyst.

@@

Meade and Wright (\protect\hyperlink{ref-meade2012solving}{2012}) compared the most commonly used AI methods and unequivocally recommended the all-others-as-anchors (AOAA) method. AOAA tests each item for DIF one at a time using all of the other items as anchors. For example, when testing the first item for DIF, all of the other items are used as anchors. This is done using a LRT that compares the baseline model, where all item parameters are fixed across groups, to the flexible model, where the parameters of the tested item are freed across groups (Thissen, Steinberg, and Wainer \protect\hyperlink{ref-thissen1993detection}{1993}). Then, when testing the second item for DIF, once again all of the other items (including the first item) are used as anchors, and so on. The items for which the flexible model outperforms the baseline model (typically based on a \(\chi^2\) test) are identified as having DIF, and the rest of the items become anchor items. AOAA is implemented in the mirt R package, and is called by passing scheme = \enquote{drop} to the DIF function (drop refers to dropping a single constraint when moving from the constrained to the flexible model).

Implicit in the use of AOAA is the assumption that all items not being tested do not exhibit DIF, which is, of course, counter to the underlying rationale for the undertaking. On a practical level, it is thought that AOAA will perform well if a small minority of items have DIF or the DIF is balanced such that some items are biased against the focal group, while others are biased against the reference group.

Researchers have noticed the circular logic of AOAA, but have mostly described it indirectly by pointing out inflated Type I errors in simulation studies (Stark, Chernyshenko, and Drasgow \protect\hyperlink{ref-stark2006detecting}{2006}). A simple thought experiment illustrates how AOAA fails: Imagine a three item test with a sufficiently large number of students where the first item has DIF, and the other two do not. Using AOAA, all items test positive for DIF. The last two items incorrectly test positive because including the first item in the anchor set causes the group ability difference to be misestimated. This phenomenon of items with real DIF inducing the appearance of DIF in other items was only indirectly discussed in the literature until Andrich and Hagquist (\protect\hyperlink{ref-andrich2012real}{2012}) coined the term \enquote{artificial DIF}, which is a significant problem in applications of AI methods as discussed below.

@@

One way to attempt to counter artificial DIF is with purification through iterative anchor selection. For example, Drasgow (\protect\hyperlink{ref-drasgow1987study}{1987}) started with AOAA, removed items displaying DIF from the anchor set, then repeated the process iteratively---with items that have been removed from the anchor set allowed to have free parameters across groups in both the baseline and flexible model---until no more items tested positively. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) named this technique Iterative-backward-AOAA with \enquote{backward} (as in reverse) referring to beginning with the assumption that all items are DIF-free. We find it clearer to refer to this method as all-others-as-anchors-all-significiant (AOAA-AS). Appending \enquote{all-significant} indicates that anchor selection is made iteratively with all items that test positive for DIF being removed from the anchor set. AOAA-AS is implemented in the mirt R package, and is called by passing scheme = \enquote{drop\_sequential} to the DIF function.

We argue that AOAA-AS, while a potential improvement, doesn't solve the fundamental problem of AOAA: What does one do when all items test positive for DIF? With a sufficient sample size and at least one item with DIF, this will necessarily be the case. In our thought experiment, we get the same result with AOAA-AS as we did with AOAA: All items test positive for DIF, and there are no anchor items. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) encountered precisely this problem in their simulation study and chose to select a single anchor item randomly. Woods (\protect\hyperlink{ref-woods2009empirical}{2009}) suggested a more straightforward, one-step method: Use AOAA and select the four items that exhibit the least amount of DIF as anchor items. It's unclear if one should proceed if those four items also test positive for DIF.

@@

We propose an extension of these methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), which, to our knowledge (and surprise), has not previously been explicitly proposed. AOAA-OAT is inspired by Hagquist and Andrich (\protect\hyperlink{ref-hagquist2017recent}{2017}), who assert that \enquote{items showing DIF initially should not be resolved simultaneously but sequentially} {[}p.~6{]}. Like AOAA-AS, AOAA-OAT starts with AOAA, but only the single item exhibiting the most DIF, based on the \(\chi^2\) test statistic, is removed from the anchor set. The process continues iteratively until no new items display DIF. AOAA-OAT and AOAA-AS are similar in that they are both iterative; the difference is that AOAA-OAT takes the more conservative approach of removing only \emph{one} item in each iteration as opposed to \emph{all} items that test positive for DIF. As a result, we believe that AOAA-OAT is less likely to be \enquote{fooled} by artifical DIF. AOAA-OAT is not currently implemented in the R package mirt.

Applying AOAA-OAT to our thought experiment demonstrates its effectiveness. The initial AOAA removes the real DIF item from the anchor set because it exhibits the most DIF (i.e., it is the least homogeneous in the sense that the group difference in performance is most different from the average group difference in performance on the other two items). In the next step, both of the other items test negative for DIF, and we arrive at the correct conclusion. To work, AOAA-OAT has two requirements: First, that at least two items do not have DIF, and second, that the set of items without DIF are more homogeneous than other sets of items.

@@

Bechger and Maris (\protect\hyperlink{ref-bechger2015statistical}{2015}) proposed selecting anchor items by identifying clusters of items that function similarly and then choosing one of those clusters to be the \enquote{anchor cluster}. They pointed out that one way around the unidentifiability issue is to consider only relative item parameters. For each group, the relative easinesses for each pair of items can be stored in the (square) matrix \(\mathbf{R}^{\text{ref}}\) with entries \({R_{xy}}^{\text{ref}} = {b_x}^{\text{ref}} - {b_y}^{\text{ref}}\), where \(x\) and \(y\) both index the items. A similar matrix, \(\mathbf{R}^{\text{foc}}\), can be formed. The ultimate matrix of interest is \(\Delta \mathbf{R} \equiv \mathbf{R}^{\text{ref}}-\mathbf{R}^{\text{foc}}\) which is the \enquote{differences between groups in the pairwise differences in (easiness) between items} {[}p.~323{]}.
